\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage[letterpaper,margin=.5in]{geometry}
\usepackage{url}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
\usepackage[scaled]{helvet}
\renewcommand{\familydefault}{\sfdefault}
\pagenumbering{gobble}

\title{Analysis of Somatic Mutations In Vegetatively Reproducing Grapes}
\date{}
% \date{\today}
% \author{Adam Orr}

\begin{document}
\maketitle

% 250 words

% The abstract should include:
%  * a brief background of the project;
%  * specific aims, objectives, or hypotheses;
%  * the significance of the proposed research and relevance to public health;
%  * the unique features and innovation of the project;
%  * the methodology (action steps) to be used;
%  * expected results; and
%  * description of how your results will affect other research areas.
% Suggestions
%  * Be complete, but brief.
%  * Use all the space allotted.
%  * Avoid describing past accomplishments and the use of the first person.
%  * Write the abstract last so that it reflects the entire application.



\section{Abstract}
Cancer is a devastating disease caused by mutations in somatic cells that cause those cells to rapidly proliferate and evade the immune system. Despite this fact, little is known about somatic mutations themselves. Of particular interest are the differences between mutation rates and profiles in eukaryotes that reproduce sexually and those that reproduce by fragmentation of the soma, similar to the mode of replication of cancer cells.
\textbf{Aim 1} of this project is to test the hypothesis that the somatic mutation rate of a population of outcrossing grapes will be lower than the somatic mutation rate from a population of grapes that solely reproduce vegetatively.
This will be achieved by whole genome sequencing three leaves in triplicate from grapes from a population of wild grapes matched with those from a population of domestic, vegetatively reproducing grapes.
\textbf{Aim 2} of this project is to test the hypothesis that the somatic mutations in vegetatively reproducing grapes are negatively correlated with chromatin accessibility, while this correlation is less prominent in outcrossing grapes.
This hypothesis will be tested by applying an ATAC-seq protocol to samples collected similarly to those in Aim 1. Mutations detected in the ATAC-Seq data as well as the data collected in Aim 1 will be used to determine the genomic locations enriched and depleted in somatic mutations.
This work will improve the understanding of how somatic mutations are generated and persist, and has broad impacts in the fields of agriculture, developmental biology, and cancer biology.




\end{document}